A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-17 podc_paper/introduction.tex
working on related work

A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues.
As well as being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
Their design uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{list-some}.

This problem occurs more broadly in other data structures that keep track of a set of elements, as observed by Ruppert \cite{Rup16}.
For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
have a term in their step complexity
that depends linearly on the contention $c$, i.e., 
the number of processes that access the data structure concurrently (which can be as high
as $p$ in the worst case).
Attiya and Fouren \cite{DBLP:conf/opodis/AttiyaF17} proved 
that amortized $\Omega(c)$ steps per operation are indeed necessary
for CAS-based implementations of lock-free bag data structures, which allow processes
to insert elements and remove an arbitrary element (chosen non-deterministically).
Since a queue trivially implements a bag, this lower bound also holds for queues.
At first glance, this would seem to settle the question of the step complexity of lock-free queues.
However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
Thus, the lower bound could be stated more precisely as an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.

In this paper, we exploit this loophole to show that it is in fact possible to implement a linearizable, lock-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first to have step complexity that is polylogarithmic in $p$ and in $q$, the number of elements
stored in the queue.
For simplicity of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS instructions and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use an unbounded number of CAS instructions in the worst case
and $\Omega(p)$ CAS instructions in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one Java provides.

>>>>

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
IT uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch-and-increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

Many concurrent data structures that keep track of a set of elements also have an $\Omega(p)$ term in their step complexity, as observed by Ruppert \cite{Rup16}.
For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
have an $\Omega(c)$ term in their step complexity, where $c$ represents contention,
the number of processes that access the data structure concurrently (which can be $p$ in the worst case).
Attiya and Fouren \cite{DBLP:conf/opodis/AttiyaF17} proved 
that amortized $\Omega(c)$ steps per operation are indeed necessary
for CAS-based implementations of lock-free bag data structures, which provide operations
to insert an element or remove an arbitrary element (chosen non-deterministically).
Since a queue trivially implements a bag, this lower bound also holds for queues.
At first glance, this would seem to settle the question of the step complexity of lock-free queues.
However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
Thus, the lower bound could be stated more precisely as an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.

In this paper, we exploit this loophole.  We show that it is in fact possible to implement a linearizable, lock-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first to have step complexity that is polylogarithmic in $p$ and in $q$, the number of elements
stored in the queue.
For simplicity of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS instructions and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use an unbounded number of CAS instructions in the worst case
and $\Omega(p)$ CAS instructions in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one Java provides.

### 2022-12-17 podc_paper/related.tex
working on related work

A number of papers have described ways to reduce contention in the MS-queue.
Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its enqueued value directly
to a concurrent dequeue when the queue is empty or when concurrent operations 
can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
attempts to reduce contention by grouping concurrent enqueues into baskets.
An enqueue that fails its CAS is put in the basket with the eqnueue that succeeded.
Enqueues within a basket determine their order among themselves without having to access the back of the queue.
However, if $p$ concurrent enqueues are in the same basket
the CAS retry problem occurs when they order themselves using CAS instructions.
Both modifications still have $\Omega(p)$ amortized step complexity.

@@

-----

@@

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10}
introduced a new algorithm using a linked list of arrays. The queue is
stored in a shared array where head and tail pointers point to the
current elements in the queue. When the array is full, an empty array
is linked to the array and tail pointers are updated. A global head
points to the array containing the first element in the queue, and
each process has a local head index that points to the first element
in that array. Global tail and local tail pointers are similar. A
process updates the position of the pointers after it does an
operation. One process might go to sleep before setting the pointers,
so the pointers might be behind their real places. They mention how to
scan the arrays to update pointers while doing an operation. A process
writes an element in the location head by a \nf{CAS} instruction, so
if $p$ processes try to enqueue simultaneously, the amortized step
(and \nf{CAS}) complexity remains $\Omega(p)$. Their design is
lock-free but not wait-free.

Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a
wait-free queue that uses the fast-path slow-path methodology
introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}. Their
work is based on a circular queue using bounded memory. When a process
wishes to do an enqueue or a dequeue, it starts two paths. The fast
path  ensures good performance while the slow path ensures
termination. They show that these two paths do not affect each other
and the queue remains consistent. If a process makes no progress,
other processes help its slow path to finish. The helping phase
suffers from the CAS Retry Problem because processes compete in a
\nf{CAS} loop to decide which succeeds to help. Because of this, the
amortized complexity cannot be better than $\Omega(p)$.

The CAS Retry Problem is not limited to list-based queues; array-based
queues also share
it~\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
Our motivation is to overcome this problem and present a wait-free
sublinear queue.

\here{Check:
John Giacomoni et al. FastForward
https://dl-acm-org.ezproxy.library.yorku.ca/doi/10.1145/1345206.1345215
}

\paragraph{Other Primitives and Restricted Queues}

David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap primitive instructions and takes constant time per operation, but
uses unbounded memory.  A modification to use bounded space
is described, but it increases the time per operation to $\Omega(p)$.

Jayanti and Petrovic introduced a wait-free poly-logarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it only works for a single dequeuer. 
Our implementation uses their idea of having
a tournament tree among processes to agree on the linearization of
operations.

\here{new result I found:}
Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
called half-increment and half-max, which can be viewed as particular kinds of
double-word read-modify-write operations.
They use this as evidence that their primitives can be more efficient than CAS
since previous CAS-based queues all required $\Omega(p)$ step complexity.
Our new implementation counters this argument.

\subsection{Universal Constructions and Other Poly-log Time Data Structures}
A \textit{universal construction} is an algorithm that can implement a
shared version of any given sequential object. The first universal
construction was introduced by
Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent
queue using a universal construction. Jayanti proved an $\Omega(\log
p)$ lower bound on the worst-case shared-access time complexity of
$p$-process universal
constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that
the universal construction by Afek, Dauber, and
Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$
worst-case step complexity, using atomic access to $\Omega(p \log
p)$-bit words. Chandra, Jayanti and Tan introduced a semi-universal
construction that achieves $\textsc{O}(\log^2 p)$ shared
accesses~\cite{DBLP:conf/podc/ChandraJT98}. However, their algorithm
cannot be used to create a queue. We mention a non-practical universal
construction with a poly-log number of \nf{CAS} instructions in the
last paragraph of page 13.

@@

>>>>

A number of papers have described ways to reduce contention in the MS-queue.
Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its enqueued value directly
to a concurrent dequeue when the queue is empty or when concurrent operations 
can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
attempts to reduce contention by grouping concurrent enqueues into baskets.
An enqueue that fails its CAS is put in the basket with the enqueue that succeeded.
Enqueues within a basket determine their order among themselves without having to access the back of the queue.
However, if $p$ concurrent enqueues are in the same basket
the CAS retry problem occurs when they order themselves using CAS instructions.
Both modifications still have $\Omega(p)$ amortized step complexity.

@@

\paragraph{Array-Based Queues}
Arrays can be used to implement queues with bounded capacity, with indices
for the front and back elements updated using CAS instructions 
\cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.
Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} avoid
the capacity constraint by using a linked list of arrays.
These solutions also suffer from the CAS retry problem, with $\Omega(p)$ steps per operation.

Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch-and-increment instructions.
If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
the enqueues fall back on using a CAS to add a new array to the linked list, 
and the CAS retry problem reappears.
% I looked at the CRQ algorithm to implement each of the circular arrays.
% Here is an execution that causes livelock for 3 processes
% enqueue(A) reads t=0 and increments tail to 1
% enqueue(B) reads t=1 and increments tail to 2
% (*) dequeue reads h=0 and increments head to 1
% dequeue's CAS succeds changing array[0] to (1,R,bottom)
% dequeue reads t=2 at line 53
% enqueue(A) fails CAS on array[0]
% dequeue reads h=1 and increments head to 2
% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
% enqueue(B) fails CAS on array[1]
% enqueue(A) reads t=2 and increments tail to 3
% enqueue(B) reads t=3 and increments tail to 4
% dequeue reads t=4 at line 53
% go back to (*) and repeat this sequence
Similarly,
other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} used the fast-path slow-path methodology of Kogan and Petrank~\cite{10.1145/2370036.2145835}.
The fast path has good performance and the slow path guarantees
termination.
In worst-case executions, where processes must execute the slow path,
these implementations also take $\Omega(p)$ steps per operation, 
due either to the CAS retry problem or helping mechanisms.

%Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a
%wait-free queue that uses the fast-path slow-path methodology. Their
%work is based on a circular queue using bounded memory. When a process
%wishes to do an enqueue or a dequeue, it starts two paths. 
% They show that these two paths do not affect each other
%and the queue remains consistent. If a process makes no progress,
%other processes help its slow path to finish. The helping phase
%suffers from the CAS Retry Problem because processes compete in a
%\nf{CAS} loop to decide which succeeds to help. Because of this, the
%amortized complexity cannot be better than $\Omega(p)$.

@@

\paragraph{Universal Constructions and Other Poly-log Time Data Structures}
A \textit{universal construction} is an algorithm that can implement a
shared version of any given sequential object. The first universal
construction was introduced by
Herlihy~\cite{10.1145/114005.102808}. We can implement a concurrent
queue using a universal construction. Jayanti proved an $\Omega(\log
p)$ lower bound on the worst-case shared-access time complexity of
$p$-process universal
constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also mentions that
the universal construction by Afek, Dauber, and
Touitou~\cite{DBLP:conf/stoc/AfekDT95} can be modified to $O(\log p)$
worst-case step complexity, using atomic access to $\Omega(p \log
p)$-bit words. Chandra, Jayanti and Tan introduced a semi-universal
construction that achieves $\textsc{O}(\log^2 p)$ shared
accesses~\cite{DBLP:conf/podc/ChandraJT98}. However, their algorithm
cannot be used to create a queue. We mention a non-practical universal
construction with a poly-log number of \nf{CAS} instructions in the
last paragraph of page 13.

Need to discuss Fatourou and Kallimanis TCS 55(3)
Fatourou and Kallimanis \cite{FK14} gave a universal construction based on fetch-and-add and CAS
and used it to implement a queue....

@@

\paragraph{Restricted Queues}
David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap  instructions and takes $O(1)$ steps per operation, but
uses unbounded memory.  Bounding the space increases the steps per operation to $\Omega(p)$.

Jayanti and Petrovic introduced a wait-free poly-logarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it works only for a single dequeuer. 
Our implementation uses their idea of having
a tournament tree among processes to agree on the linearization of
operations.

\here{Should we mention https://arxiv.org/abs/2103.11926 , which seems to have a queue for 2 enqueuers and 2 dequeuers ?  May not be worth mentioning because then $p$ is a constant, so complexity in terms of $p$ doesn't make sense.}

\paragraph{Other Primitives}
Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
called half-increment and half-max, which can be viewed as particular kinds of
double-word read-modify-write operations.
They use this as evidence that their primitives can be more efficient than CAS
since previous CAS-based queues all required $\Omega(p)$ step complexity.
Our new implementation counters this argument.
